### 2021-10-29 text/outline/draft.tex
update abstract

\begin{abstract}
CRISPR genome engineering and single-cell sequencing have transformed biological discovery. Single-cell CRISPR screens unite these two technologies, linking genetic perturbations in individual cells to changes in gene expression and illuminating regulatory networks underlying diseases. In this work we study single-cell CRISPR screens from a statistical perspective. First, we demonstrate on real data that a standard method for estimation and inference in single-cell CRISPR screens — “thresholded regression” — exhibits attenuation bias and a bias-variance tradeoff as a function of an intrinsic tuning parameter. We recover these phenomena in precise theoretical terms in an idealized Gaussian setting. Next, we introduce GLM-EIV (“generalized linear model with errors-in-variables”), a new method for single-cell CRISPR screen analysis. GLM-EIV generalizes the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed, overcoming limitations of thresholded regression. We develop a computational infrastructure to deploy GLM-EIV across hundreds or thousands of processors on clouds (e.g., Microsoft Azure) and high-performance clusters. Leveraging this infrastructure, we apply GLM-EIV to analyze two recent, large-scale, single-cell CRISPR screen datasets, yielding new biological insights.
\end{abstract}

@@

The most promising biotechnology for solving this challenge are high MOI single-cell CRISPR screens. High MOI single-cell CRISPR screens combine CRISPR interference (CRISPRi) -- a version of CRISPR that \textit{represses} a targeted region of the genome, leaving the primary DNA sequence intact  -- with single-cell sequencing. The experimental protocol is as follows. First, the scientist develops a library of several hundred to several thousand CRISPRi perturbations, each designed to target a candidate enhancer for repression. The scientist then cultures tens or hundreds of thousands of cells and delivers the CRISPRi perturbations to the cells. The perturbations assort into the cells randomly, with each cell receiving on average 10-40 distinct perturbations. Conversely, a given perturbation enters fewer than 0.1-2\% of cells.

After waiting several days for CRISPRi to take effect, the scientist profiles each cell's transcriptome (i.e., its gene expressions) and the set of perturbations that it received. Finally, the scientist conducts perturbation-to-gene association analyses. Figure \ref{analysis_challenges}a depicts this process schematically, with colored bars (blue, red, and purple) representing distinct perturbations. For a given perturbation (e.g., the perturbation represented in blue), the scientist partitions the cells into two groups: those that received the perturbation (top) and those that did not (bottom). Next, for a given gene, the scientist runs a differential expression analysis across the two groups of cells, producing an estimate for the magnitude of gene expression change in response to the perturbation. If the estimated change in expression is large, the scientist can conclude that the enhancer \textit{targeted} by the perturbation exerts a strong regulatory effect on the gene. This procedure is repeated for a large set of preselected perturbation-gene pairs. % If the differential expression analyses produce $p$-values, the $p$-values are subjected to a multiple hypothesis testing correction procedure.

\subsection{Analysis challenges and data}
High MOI single-cell CRISPR screens present several statistical challenges. We take a parametric approach in this work, and so we highlight challenges that are especially relevant to modeling the data generating process. Throughout, we consider a single perturbation-gene pair. First, the ``treatment'' variable -- i.e., the presence or absence of a perturbation -- is unobservable. Perturbed cells transcribe molecules called  \textit{guide RNAs} (or \textit{gRNAs}) that serve as indirect signals of perturbation presence. We can leverage the gRNA data to infer whether a given cell has been perturbed. Using these imputed perturbation assignments, we can estimate the effect of the perturbation on gene expression (Figure \ref{analysis_challenges}b). Second, ``technical factors'' -- sources of variation that are experimental rather than biological in origin -- impact both gene expression and gRNA expression (Figure \ref{analysis_challenges}b).  We must account for these technical factors in constructing models for gene and gRNA expression. Third, the gene and gRNA data are sparse, discrete counts. Therefore, classical statistical approaches that assume Gaussianity or homoscedasticity are inapplicable. Finally, and most subtly, sequenced gRNAs sometimes are mapped to unperturbed cells. This phenomenon, which we call the ``background read'' phenomenon, results from errors in the sequencing process \cite{Replogle2020}. The marginal distribution of the gRNA count data is best conceptualized as a mixture model (Figure \ref{analysis_challenges}c; Gaussian distributions used for illustration purposes only). Unperturbed and perturbed cells both exhibit nonzero gRNA count distributions, but the gRNA count distribution for perturbed cells is greater. Figure \ref{analysis_challenges}d shows example data for four (of possibly tens or hundreds of thousands of) cells. The analysis objective is to leverage the observed data -- i.e., the gene expressions and gRNA counts -- to estimate the effect of the (latent) perturbation on gene expression, accounting for the technical factors.

>>>>

\begin{abstract}
CRISPR genome engineering and single-cell sequencing have transformed biological discovery. Single-cell CRISPR screens unite these two technologies, linking genetic perturbations in individual cells to changes in gene expression. Single-cell CRISPR screens promise to illuminate the regulatory wiring underlying diseases but present substantial statistical challenges. We demonstrate on real data that a standard method for estimation and inference in single-cell CRISPR screens — “thresholded regression” — exhibits attenuation bias and a bias-variance tradeoff as a function of an intrinsic tuning parameter. We recover these phenomena in precise theoretical terms in an idealized Gaussian setting. Next, we introduce GLM-EIV (“generalized linear model with errors-in-variables”), a new method for single-cell CRISPR screen analysis. GLM-EIV generalizes the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed, overcoming limitations of thresholded regression. We develop a computational infrastructure to deploy GLM-EIV across hundreds or thousands of processors on clouds (e.g., Microsoft Azure) and high-performance clusters. Leveraging this infrastructure, we apply GLM-EIV to analyze two recent, large-scale, single-cell CRISPR screen datasets, yielding new biological insights.
\end{abstract}

@@

The most promising biotechnology for solving this challenge are high MOI single-cell CRISPR screens. High MOI single-cell CRISPR screens combine CRISPR interference (CRISPRi) -- a version of CRISPR that \textit{represses} a targeted region of the genome, leaving the primary DNA sequence intact  -- with single-cell sequencing. The experimental protocol is as follows. First, the scientist develops a library of several hundred to several thousand CRISPRi perturbations, each designed to target a candidate enhancer for repression. The scientist then cultures tens or hundreds of thousands of cells and delivers the CRISPRi perturbations to the cells. The perturbations assort into the cells randomly, with each cell receiving on average 10-40 distinct perturbations. Conversely, a given perturbation enters about 0.1-2\% of cells.

After waiting several days for CRISPRi to take effect, the scientist profiles each cell's transcriptome (i.e., its gene expressions) and the set of perturbations that it received. Finally, the scientist conducts perturbation-to-gene association analyses. Figure \ref{analysis_challenges}a depicts this process schematically, with colored bars (blue, red, and purple) representing distinct perturbations. For a given perturbation (e.g., the perturbation represented in blue), the scientist partitions the cells into two groups: those that received the perturbation (top) and those that did not (bottom). Next, for a given gene, the scientist runs a differential expression analysis across the two groups of cells, producing an estimate for the extent to which the gene's expression has changed in response to the perturbation. If the estimated change in expression is large, the scientist can conclude that the enhancer \textit{targeted} by the perturbation exerts a strong regulatory effect on the gene. This procedure is repeated for a large set of preselected perturbation-gene pairs. % If the differential expression analyses produce $p$-values, the $p$-values are subjected to a multiple hypothesis testing correction procedure.

\subsection{Analysis challenges}
High MOI single-cell CRISPR screens present several statistical challenges. We take a parametric approach in this work, and so we highlight challenges that are especially relevant to modeling the data generating process. Throughout, we consider a single perturbation-gene pair. First, the ``treatment'' variable -- i.e., the presence or absence of a perturbation -- is unobservable. Perturbed cells transcribe molecules called  \textit{guide RNAs} (or \textit{gRNAs}) that serve as indirect signals of perturbation presence. We can leverage the gRNA data to infer whether a given cell has been perturbed. Using these imputed perturbation assignments, we can estimate the effect of the perturbation on gene expression (Figure \ref{analysis_challenges}b). Second, ``technical factors'' -- sources of variation that are experimental rather than biological in origin -- impact both gene expression and gRNA expression (Figure \ref{analysis_challenges}b).  We must account for these technical factors in constructing models for gene and gRNA expression. Third, the gene and gRNA data are sparse, discrete counts. Therefore, classical statistical approaches that assume Gaussianity or homoscedasticity are inapplicable. Finally, and most subtly, sequenced gRNAs sometimes are mapped to unperturbed cells. This phenomenon, which we call the ``background read'' phenomenon, results from errors in the sequencing process \cite{Replogle2020}. The marginal distribution of the gRNA count data is best conceptualized as a mixture model (Figure \ref{analysis_challenges}c; Gaussian distributions used for illustration purposes only). Unperturbed and perturbed cells both exhibit nonzero gRNA count distributions, but the gRNA count distribution for perturbed cells is greater. Figure \ref{analysis_challenges}d shows example data for four (of possibly tens or hundreds of thousands of) cells. The analysis objective is to leverage the observed data -- i.e., the gene expressions and gRNA counts -- to estimate the effect of the (latent) perturbation on gene expression, accounting for the technical factors.
